
---
title: "Part 1 - Basic Inferential Data Analysis"
author: "Peter Maneykowski"
date: "4/15/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load `ToothGrowth` data set and do initial transformations.
```{r}
library(datasets); library(ggplot2)
data("ToothGrowth")
toothGrowth <- ToothGrowth 
toothGrowth$dose <- as.factor(toothGrowth$dose)  # not loaded as categorical variable
```
**Basic exploratory data analyses.**

Top of data set.
```{r}
head(toothGrowth)
```
Bottom of data set.
```{r}
tail(toothGrowth)
```
Unique values of `supp`: supplement type, either vitamin C (`VC`) or orange
juice (`OJ`)
```{r}
unique(toothGrowth$supp)
```
Unique values of `dose`: dosage in milligrams (of vitamin C) per day. `dose` 
is delivered as either a vitamin C supplement or orange juice.
```{r}
unique(toothGrowth$dose)
```
A sorted vector of unique values of `len`: length of teeth (in arbitrary
units)
```{r}
sort(unique(toothGrowth$len))
```

**Basic summary** of the data set with an accompanying boxplot demonstrating
the same relationship as the `summary` function.
```{r}
summary(toothGrowth)
```
```{r}
ggplot(toothGrowth, aes(x=dose, y=len, fill=dose)) + 
  geom_boxplot() +
  ggtitle("Fig. 1: dispersion of tooth growth by dose") +
  xlab("dose in mg") +
  ylab("tooth length")
```

**Confidence intervals and/or hypothesis tests to compare tooth growth by supp
and dose**:

**Hypothesis 1**: Orance juice and vitamin C produce the same tooth growth
across the entire data set.
```{r}
h1<-t.test(len ~ supp, data = toothGrowth)
h1$conf.int
h1$p.value
```
`conf.int` includes `0` and `p.value` is greater than the threshold of `0.05`,
so the null hypothesis that we created cannot be rejected.

**Hypothesis 2**: Orance juice and vitamin C produce the same tooth growth for
a dosage of `0.5`mg/day.
```{r}
h2<-t.test(len ~ supp, data = subset(toothGrowth, dose == 0.5))
h2$conf.int
h2$p.value
```
`conf.int` does not include `0`, and `p.value` is below our threshold of
`0.05`, so the null hypothesis can be rejected. The alternate hypothesis that
at a dosage of `0.5`mg/day orange juice produces more tooth growth than
vitamin C is accepted.

**Hypothesis 3**: Oranje juice and vitamin C produce the same tooth growth for
a dosage of `1.0`mg/day.
```{r}
h3<-t.test(len ~ supp, data = subset(toothGrowth, dose == 1))
h3$conf.int
h3$p.value

```
`conf.int` does not include `0`, and `p.value` is below our threshold of
`0.05`, so the null hypothesis can be rejected. The alternate hypothesis that
at a dosage of `1.0`mg/day orange juice produces more tooth growth than
vitamin C is accepted.

**Hypothesis 4**: Oranje juice and vitamin C produce the same tooth growth for
a dosage of `2.0`mg/day.
```{r}
h4<-t.test(len ~ supp, data = subset(toothGrowth, dose == 2))
h4$conf.int
h4$p.value
```
`conf.int` does not include `0` and `p.value` is greater than the threshold of
`0.05`, so the null hypothesis that we created cannot be rejected.

**Conclusions and Assumptions**
By using `t.test` for hypothesis testing, we may conclude that for dosages of
`0.5`mg/day and `1.0`mg/day orange juice produces more tooth growth than
vitamin C. Further, we may conclude that for a dosage of `2.0`mg/day orange
juice and vitamin C produce the same amount of tooth growth. Finally, when
looking at the entire data set, we cannot conclude that using oranje juice is
more effective than using vitamin C. The assumptions used for testing were that
there is a normal distribution of tooth lengths, and that there were no other
factors that affect tooth length.
