# Exploration of NOAA Storm Database and Analysis of Severe Weather Events

<!--

Your data analysis must address the following questions:

1. Across the United States, which types of events (as indicated in the
   `EVTYPE` variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic
   consequences?

-->

## Synopsis

<!-- at most 10 complete sentences -->

By using the [NOAA Storm Database][1] ([csv][2]) and corresponding 
[documentation][3] we are able to better understand the effects of severe
weather. We do a brief analysis grouping and plotting severe weather events
by category and resulting quantitative fields.

It is useful to understand how different factors relate to one and other as it
may be helpful in preventing or preparing for such disasters in the future.

[1]: https://www.ncdc.noaa.gov/stormevents/
[2]: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
[3]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf


## Data Processing

Load required libararies

```{r}
library(ggplot2)
```


```{r}
datapath = "data/StormData.csv.bz2"
dir.create(dirname(datapath), showWarnings = FALSE)
if (! file.exists(datapath))
    download.file(
        "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", datapath)
```

The data should be in the same directory as the script. Read in the data set
(902,297 observations) with typing on initial load to save time.

```{r}
# Initial data frame
df <- read.csv(datapath, header = TRUE, sep = ",",
               colClasses = c("numeric", "character", "character", "character",
                              "numeric", "character", "character", "character",
                              "numeric", "character", "character", "character",
                              "character", "numeric", "character", "character",
                              "character", "character", "numeric", "numeric",
                              "character", "numeric", "numeric", "numeric",
                              "numeric", "character", "numeric", "character",
                              "character", "character", "character", "numeric",
                              "numeric", "numeric", "numeric", "character",
                              "numeric"))
```

For this study, we are only interested in the following four categories. This
brings the row count of the data frame down to 254,633 observations.

```{r}
# Small (filtered) data frame
sf <- df[df$FATALITIES > 0
         | df$INJURIES > 0
         | df$PROPDMG > 0
         | df$CROPDMG > 0,]
```

Now, we would like to group the `EVTYPE` (event type) into a categorial variable. There are
six large categories: `Convection`, `Extreme Temperatures`, `Flood`, `Marine`,
`Cyclone`, and `Winter`. I have devised a regular expression to cover all the
major possibilities for the six categories. In addition to those, `Other`
covers everything else.

```{r}
names <- c("Convection", "Extreme Temperatures", "Flood", "Marine",
           "Cyclone", "Winter")
expressions <- c(
    paste("?.*lighting|lightn|ligntning|tornado|torndao|wind|tstm|thunder|wnd",
          "|hail|turbulence|gustnado?.*", sep = ""),
    "?.*cold|heat|warm|temperature|thermia?.*",
    "?.*flood|rising|stream fld?.*",
    "?.*coastal|tsunami|current|swells|tide|wave|seas|surf|high water?.*",
    "?.*cyclone|hurricane|typhoon?.*",
    "?.*wint|ice|avalan|sleet|snow|freez|blizz|frost|glaze|mixed?.*")
for (i in 1:length(names)) {
    mask <- grepl(expressions[i], sf$EVTYPE, ignore.case = TRUE)
    sf[mask, "category"] = names[i]
}
mask <- is.na(sf$category)
sf[mask, "category"] = "Other"
```

Below is a code snippet to convert the `propertydamageEXP` and `cropdamageEXP` 
fields to numerical variables. If the field contains the pattern `h|H`, it
recieves a value of `100`, `k|K` recieves `1000`, and so on. These values are
specified in the corresponding documentation.

```{r}
sf$propertydamageEXP = 1
patterns <- c("h|H", "k|K", "m|M", "b|B")
values <- c(100, 1000, 1e+06, 1e+09)
for (i in 1:length(values)) {
    mask <- grepl(patterns[i], sf$PROPDMGEXP, ignore.case = TRUE)
    sf[mask, "propertydamageEXP"] = values[i]
}
sf$propertydamage = sf$PROPDMG * sf$propertydamageEXP

sf$cropdamageEXP = 1
patterns <- c("k|K", "m|M", "b|B")
values <- c(100, 1e+06, 1e+09)
for (i in 1:length(values)) {
    mask <- grepl(patterns[i], sf$CROPDMGEXP, ignore.case = TRUE)
    sf[mask, "cropdamageEXP"] = values[i]
}
sf$cropdamage = sf$CROPDMG * sf$cropdamageEXP
```


## Results

### Weather events that cause the largest number of incidents

Create an `incidents` column that is comprised of the number of fatalities and
injuries. This will be used to discover which severe weather events have the
highest number of incidents.

```{r}
sf$incidents = sf$FATALITIES + sf$INJURIES
```

Create a frame that shows the name of the event with corresponding number of
incidents. It shows that `Convection` events have the greatest number of 
incidents.

```{r}
# Incident data
inc <- aggregate(list(incidents = sf$incidents),
                 by = list(event = sf$category),
                 FUN = sum, na.rm = TRUE)
```

A corresponding plot showing the name of event and number of incidents.

```{r}
inc$event <- reorder(inc$event, -inc$incidents)

ggplot(inc, aes(y = incidents)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    geom_bar(aes(x = event), data = inc, stat = "identity") +
    ggtitle("Fatalities and Injuries") +
    xlab("Event Category") +
    ylab("No. of Fatalities and Injuries")
```

### Weather events that cause the largest amount of damage

Create a `damage` field that contains damage to property as well as crops. This
is so that we can find out which events cause the most dollar damage. This 
column is in billions of dollars.

```{r}
sf$damage = ((sf$propertydamage + sf$cropdamage) / 1e+09)
```

Create a frame that has name of event corresponding to the dollar amount of 
damage (in billions). The frame shows that `Flood` results in the greatest 
amount of dollar damage.

```{r}
# Damage data frame
dmg <- aggregate(list(damage = sf$damage),
                 by = list(event = sf$category),
                 FUN = sum, na.rm = TRUE)
```

A corresponding plot showing the name of event and billions of dollars in 
damage.

```{r}
dmg$event <- reorder(dmg$event, -dmg$damage)

ggplot(dmg, aes(y = damage)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    geom_bar(aes(x = event), data = dmg, stat = "identity") +
    ggtitle("Property and crop damage") +
    xlab("Event Category") +
    ylab("Amount of damage (billions of $)")
```


<!-- vim: colorcolumn=80 nowrap
-->